Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-22 ms.tex
Rewording things.

Between outbursts \citealt{Ciotti+10} find that galactic nuclei exist
in a quasi-stationary thermal stable state. We focus on this region of
parameter space calculating time-independent models, in which the
nuclear gas receives sufficient heating to render radiative cooling
negligible.  This approach allows us to systematically explore the
relevant parameter space and to derive analytic expressions that prove
useful in determining under what conditions cooling instabilities
manifest across the expected range of galaxy properties, or whether
other (non-AGN) forms of feedback can produce a prolonged state of
steady, thermally stable accretion.  Even if cooling instabilities
develop on galactic scales over longer $\sim$ Gyr timescales, a quasi
steady-state may exist between these inflow events on smaller radial
scales comparable to the sphere of influence.  We are thus motivated
to explore whether such steady flows are generally expected within the
nucleus, or whether in most galaxies the nuclear gas content is always
building towards instability.

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ usually resides near the nominal Bondi
radius).  When heating is sufficiently weak, however, the stagnation
radius may move to much larger radii or not exist at all, in which
case nearly the entire ISM of the galaxy is flowing towards the SMBH.
Here we assess under what conditions a stagnation
radius exists, a question intimately tied to that of cooling
instability. {\bf AG this section still needs work...}

@@

We fix $\beta = 2$ because our results are not overly sensitive to the properties of the gas flow on radial scales $\gtrsim r_b$. The presence of the break radius $r_{b}$ is,
however, necessary to obtain a converged steady state for some regions
of our parameter space.  We consider solutions calculated for up to
four values of $\rb$: 50 pc, 100 pc, 200 pc, and 400 pc, motivated by
the range of break radii from the \citet{LauerFaber+:2007a}
sample.\footnote{For the core galaxies the break radius follows the
  scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with
  scatter of approximately one dex.  Most the cusp galaxies have $\rb$
  between 100 and 1000 pc, and lack a clear trend with $\Mbh$.}  In
practice we calculate solutions for a range of $\rb$ only in cases
when its precise value is expected to have a significant affect on the
solution, in particular for high SMBH masses and low heating rates,
for which the stagnation radius approaches large radii. We also
neglect value of $r_b$ which would give unphysically large bulges for
a given $\Mbh$.

@@

Following \citet{Quataert:2004a} (see also
\citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}),
we calculate the density $\rho$, temperature $T$, and radial velocity
$v$ of the CNM for each galaxy model by solving the equations of
one-dimensional, time-dependent hydrodynamics,
\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial
      v}{\partial r}\right) =-\frac{\partial p}{\partial r}- \rho\frac{GM_{\rm enc}}{r^{2}} -q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial
      s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] 
, 
\label{eq:dsdt}
\end{align}
where $p$ and $s$ are the pressure and specific entropy, respectively, and $M_{\rm enc} = M_{\star}(r) + \Mbh$ is the
enclosed mass (we neglect dark matter contributions).  We adopt an ideal gas equation of state with $p = \rho kT/\mu m_p$ with $\mu = 0.62$  and $\gamma = 5/3$.

@@

The accretion rate onto the SMBH is given by the total mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& 4\pi \int_{0}^{r_{s}}q r^{2}dr = \frac{\eta M_{\star}|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    4.5 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    3.2 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have assumed $v_{w} \gg \sigma_0$ by adopting equation
(\ref{eq:stag_simple}) for $r_s$.  The resulting Eddington ratio is
given by
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    2.0 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core}, \\
    1.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$ is the
Eddington accretion rate, assuming a radiative efficiency of ten per
cent.  Note the sensitive dependence of the accretion rate on the wind
heating rate. Also, we emphasize that
equation~\eqref{eq:eddr_analytic} represents the mass inflow rate on
large scales and does not include small scale outflows. In reality, only
a fraction $\alpha$ (the standard Shakura-Sunyaev parameter--assumed
to be 0.1) will make it down to the black hole \citep{Li+2013}.

@@

%Note that by combining equations~\eqref{eq:mdot_analytic}
%and~\eqref{eq:mdot_gas}, one can also write
%\begin{align}
%\rho|_{r_{\rm s}}=\frac{3 q|_{r_{\rm s}} \tff|_{r_{\rm s}}}{2-\Gamma}.
%\label{eq:rhors2}
%\end{align}

@@

For $r\simeq r_{\rm circ} $ the gas will encounter a centrifugal
barrier, which means that much of the gas flowing in from larger
scales will be driven out in a conical outflow. The actual accretion
rate will be $\sim\alpha$ times the accretion rate on large scales
\citep{Li+2013}.

@@

\citet{Sharma+2007} calculate the radiative efficiency, $\epsilon$,
based on MHD shearing box simulations of collisionless plasmas. For
accretion rates $10^{-4} \lsim \eddr \lsim 10^{-2}$, the radiative
efficiency is $\sim 2.6\%$ {\bf AG described in more
  detail?}. Additionally, a large fraction of the $\dot{M}$ on large scales
will be driven out in outflows as the gas has some non-zero angular
momentum and a centrifugal barrier will stop it from accreting
\citet{Li+2013}. In the limit weak cooling appropriate for low
accretion rates the accretion rate on small scales will be of order
$\alpha \Mdot$. Here we adopt $\alpha=0.1$ so that our overall
efficiency is 0.26\%

@@

\begin{figure*}
\includegraphics[width=\textwidth]{miller.pdf}
\caption{\label{fig:miller} Average nuclear X-ray luminosity, $\langle
  L_X \rangle = \epsilon_X \Mdot c^2$, as a function of SMBH mass.
  Green lines show our prediction in the case of heating due to the
  average SFH for galaxies corresponding to each black hole mass
  (Fig.~\ref{fig:bh_growth}), calculated using a power-law fit for the
  relationship $\Gamma(M_{\bullet})$ derived from the
  \citet{LauerFaber+:2007a} sample.  Shown for comparison are the
  measurements ({\it black stars}) and upper limits ({\it gray 
    triangles}) for $L_X/L_{\rm edd}$ values, from the
  \citet{Miller+15} sample of early-type galaxies (a black line shows
  the best power-law fit $\langle L_X/L_{\rm edd}\rangle \propto
  M_{\bullet}^{\alpha}$, with $\alpha = -0.2$).  The left panel is
  calculated assuming a constant radiative efficiency
  $\epsilon_X=10^{-4} (\epsilon_{\rm bol}/0.1)$, while the right hand
  panel assumes $\epsilon_X= (\epsilon_{\rm
    bol}/0.1)(\Mdot/\MdotEdd)$, as predicted by ADAF models (e.g.,
  \citealt{Narayan&Yi95}).  Shown for comparison are the X-ray
  luminosities calculated for the maximum thermally-stable accretion
  rate ({\it dashed orange line}; eq.~[\ref{eq:Mdotmax}]); the SN
  Ia-regulated accretion rate ({\it dashed teal line};
  eq.~[\ref{eq:eddr_Ia}]), and the Compton heating-regulated accretion
  rate ({\it dashed pink line}; eq.~[\ref{eq:MdotCa}]), again all
  calculated for the average SFH corresponding to each SMBH mass.}
\end{figure*}

>>>>

In the this paper we focus on time-independent models, in which the
nuclear gas receives sufficient heating to render radiative cooling
negligible.  This approach allows us to systematically explore the
relevant parameter space and to derive analytic expressions that prove
useful in determining under what conditions cooling instabilities
manifest across the expected range of galaxy properties, or whether
other (non-AGN) forms of feedback can produce a prolonged state of
steady, thermally stable accretion.  Even if cooling instabilities
develop on galactic scales over longer $\sim$ Gyr timescales, a quasi
steady-state may exist between these inflow events on smaller radial
scales comparable to the sphere of influence.  We note that
\citealt{Ciotti+10} find that between outbursts gas may exist in a
quasi-steady state {\bf AG reworded}.

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ usually resides near the nominal Bondi
radius).  When heating is sufficiently weak, however, the stagnation
radius may move to much larger radii or not exist at all, in which
case nearly the entire ISM of the galaxy is flowing towards the SMBH.
Here we assess under what conditions a stagnation
radius exists, a question intimately tied to that of cooling
instability.

@@

We fix $\beta = 2$ because our results are not overly sensitive to the
properties of the gas flow on radial scales $\gtrsim r_b$ {\bf AG:
  beta=1 could be problematic for low heating rates}. The presence of
the break radius $r_{b}$ is, however, necessary to obtain a converged
steady state for some regions of our parameter space.  We consider
solutions calculated for up to four values of $\rb$: 50 pc, 100 pc,
200 pc, and 400 pc, motivated by the range of break radii from the
\citet{LauerFaber+:2007a} sample.\footnote{For the core galaxies the
  break radius follows the scaling relationship $\rb\sim 106 \,
  (\Mbheight)^{0.39}$ pc, with scatter of approximately one dex.  Most
  the cusp galaxies have $\rb$ between 100 and 1000 pc, and lack a
  clear trend with $\Mbh$.}  In practice we calculate solutions for a
range of $\rb$ only in cases when its precise value is expected to
have a significant affect on the solution, in particular for high SMBH
masses and low heating rates, for which the stagnation radius
approaches large radii. We also neglect values of $r_b$ which would
give unphysically large bulges for a given $\Mbh$.

For low mass galaxies, nuclear star clusters $\Mbh\lsim 2\times 10^8
\Msun$ may be present \citep{GrahamSpitler:2009a}.  Mass injection to
the nucleus could (for $\Mbh\lsim 2 \times 10^{8} \Msun$) therefore be
dominated by a nuclear star cluster with its own break radius on
parsec scales, i.e. much smaller than our fiducial value of $r_b \sim
100$ pc. However, for high heating rates the stagnation radius is not
sensitive to the location of the break radius, and out models could
also describe the accretion rate in the case of a nuclear star cluster.

@@

Following \citet{Quataert:2004a} (see also
\citealt{HolzerAxford:1970a,De-ColleGuillochon+:2012a,ShcherbakovWong+:2014a}),
we calculate the density $\rho$, temperature $T$, and radial velocity
$v$ of the CNM for each galaxy model by solving the equations of
one-dimensional, time-dependent hydrodynamics,

\begin{align}
  &\drhodt+\frac{1}{r^2}\frac{\partial}{\partial r}\left(\rho r^2 v\right)=q \label{eq:drhodt}\\
  &\rho \left(\frac{\partial v}{\partial t} + v\frac{\partial
      v}{\partial r}\right) =-\frac{\partial p}{\partial r}- \rho\frac{GM_{\rm enc}}{r^{2}} -q v \label{eq:dvdt}\\
  &\rho T\left(\frac{\partial s}{\partial t} + v\frac{\partial
      s}{\partial r}\right)=q\left[\ke+\kew-\gammaf \cs \right] ,
\label{eq:dsdt}
\end{align}

where $p$ and $s$ are the pressure and specific entropy, respectively,
and $M_{\rm enc} = M_{\star}(r) + \Mbh$ is the enclosed mass (we
neglect dark matter contributions).  We adopt an ideal gas equation of
state with $p = \rho kT/\mu m_p$ with $\mu = 0.62$ and $\gamma = 5/3$.

@@

The accretion rate onto the SMBH is given by the total mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& 4\pi \int_{0}^{r_{s}}q r^{2}dr = \frac{\eta M_{\star}|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    4.5 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    3.2 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have assumed $v_{w} \gg \sigma_0$ by adopting equation
(\ref{eq:stag_simple}) for $r_s$.  The resulting Eddington ratio is
given by
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    2.0 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core}, \\
    1.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$ is the
Eddington accretion rate, assuming a radiative efficiency of ten per
cent.  Note the sensitive dependence of the accretion rate on the wind
heating rate. Also, we emphasize that
equation~\eqref{eq:eddr_analytic} represents the mass inflow rate on
large scales and does not include small scale
outflows. \citep{Li+2013} suggest only a fraction $\alpha$ (the
standard Shakura-Sunyaev parameter--assumed to be 0.1) will make it
down to the black hole.

@@

@@

For $r\simeq r_{\rm circ} $ the gas will encounter a centrifugal
barrier, which means that much of the gas flowing in from larger
scales will be driven out in a conical outflow. \citep{Li+2013}
suggest that the actual accretion rate will be $\sim\alpha$ times the
accretion rate on large scales.

@@

\citet{Sharma+2007} calculate the radiative efficiency, $\epsilon$,
based on MHD shearing box simulations of collisionless plasmas. For
accretion rates $10^{-4} \lsim \eddr \lsim 10^{-2}$, the radiative
efficiency is $\epsilon_{\rm rad}\sim 2.6\%$\footnote{for $\eddr\lsim
  10^{-4} \epsilon_{\rm rad}\sim \eddr^{0.9}$. For $\eddr\gg 0.01$
  $\epsilon_{\rm rad}\sim 0.1$...}. Additionally, a large
fraction of the $\dot{M}$ on large scales will be driven out in
outflows as the gas has some non-zero angular momentum and a
centrifugal barrier will stop it from accreting \citet{Li+2013}. In
the limit weak cooling appropriate for low accretion rates the
accretion rate on small scales will be of order $\alpha \Mdot$. Here
we adopt $\alpha=0.1$ so that our overall efficiency is 0.26\%

@@

\begin{figure*}
\includegraphics[width=\textwidth]{miller.pdf}
\caption{\label{fig:miller} Average nuclear X-ray luminosity, $\langle
  L_X \rangle = \epsilon_X \Mdot c^2$, as a function of SMBH mass.
  Green lines show our prediction in the case of heating due to the
  average SFH for galaxies corresponding to each black hole mass
  (Fig.~\ref{fig:bh_growth}), calculated using a power-law fit for the
  relationship $\Gamma(M_{\bullet})$ derived from the
  \citet{LauerFaber+:2007a} sample.  Shown for comparison are the
  measurements ({\it black stars}) and upper limits ({\it gray
    triangles}) for $L_X/L_{\rm edd}$ values, from the
  \citet{Miller+15} sample of early-type galaxies (a black line shows
  the best power-law fit $\langle L_X/L_{\rm edd}\rangle \propto
  M_{\bullet}^{\alpha}$, with $\alpha = -0.2$).  The left panel is
  calculated assuming a constant radiative efficiency
  $\epsilon_X=10^{-4} (\epsilon_{\rm bol}/0.1)$, while the right hand
  panel assumes $\epsilon_X= 0.01 (\epsilon_{\rm bol}/0.1)
  (\alpha/0.1) \epsilon_{rad} (\eddr) $, where $\epsilon_{\rm
    rad}\simeq 0.026$ for $\lsim 10^{-4} \eddr\lsim 10^{-2} $, as
  motivated by shearing box simulations of collisionless plasma (see
  \citet{Sharma+2007}).  Shown for comparison are the X-ray
  luminosities calculated for the maximum thermally-stable accretion
  rate ({\it dashed orange line}; eq.~[\ref{eq:Mdotmax}]); the SN
  Ia-regulated accretion rate ({\it dashed teal line};
  eq.~[\ref{eq:eddr_Ia}]), and the Compton heating-regulated accretion
  rate ({\it dashed pink line}; eq.~[\ref{eq:MdotCa}]), again all
  calculated for the average SFH corresponding to each SMBH mass.}
\end{figure*}
